---
title: "ERBB2 Network Analysis with NetWeave"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ERBB2 Network Analysis with NetWeave}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates a complete NetWeave analysis using the bundled bladder cancer expression dataset from Sjodahl et al. (2017). We build a STRING protein-protein interaction network anchored on **ERBB2**, annotate genes with subtype expression profiles, Human Protein Atlas (HPA) protein evidence, and network centrality metrics, then visualise the results.

The same workflow can be run step-by-step (shown below) or via the `run_netweave()` convenience wrapper.

---

## Setup

```{r library}
library(NetWeave)
```

The bundled datasets are available immediately after loading the package:

```{r data}
# Expression matrix (genes x samples)
data(sjodahl_2017)

# Named subtype vector (sample IDs → subtype labels)
data(sjodahl_2017_subtypes)

# Independent expression cohort for low-expression thresholding
data(usq_expr)

# HPA reference tables
data(hpa_normal_data)
data(hpa_cancer_data)
```

---

## Step-by-Step Pipeline

### Step 1 — Retrieve pathway genes

Fetch all genes in the Reactome "Signaling by ERBB2" pathway. The function
queries the bundled `Ensembl2Reactome` object and returns a character vector
of HGNC gene symbols.

```{r step1}
erbb2_pathway <- get_pathway_genes(
  this_pathway = "Signaling by ERBB2",
  verbose      = TRUE
)
```

### Step 2 — Build and expand the STRING network

Starting from the pathway genes, `create_and_expand_network()` maps genes to
STRING protein IDs, retrieves first-degree neighbours present in the expression
data, selects the top `max_added_genes` by interaction degree, and computes
per-node centrality metrics.

Pre-downloaded STRING files are used automatically when `string_data_dir`
points to a directory containing the required `.txt.gz` flat files. A score
threshold of 900 retains only high-confidence interactions.

```{r step2}
erbb2_network <- create_and_expand_network(
  expr_data              = sjodahl_2017,
  seed_gene              = "ERBB2",
  pathway_genes          = erbb2_pathway,
  max_added_genes        = 20,
  string_score_threshold = 900,
  verbose                = TRUE,
  genes_blacklist        = NULL,
  string_data_dir        = "data/string_db/"
)

# The result is a list with two elements:
edge_data    <- erbb2_network$edge_data    # data.frame of network edges
node_metrics <- erbb2_network$node_metrics # data.frame of per-node metrics
```

### Step 3 — Add per-subtype mean expression

Appends `mean_expr_<subtype>` columns to `node_metrics`. The Sjodahl 2017
cohort contains five molecular subtypes; we specify their display order
explicitly.

```{r step3}
annotated_node_metrics <- add_subtype_expression(
  this_subtype_vector = sjodahl_2017_subtypes,
  expr_data           = sjodahl_2017,
  node_metrics        = erbb2_network$node_metrics,
  subtype_order       = c("Uro", "GU", "BaSq", "Mes", "ScNE")
)
```

### Step 4 — Flag low-expressed genes

Uses a separate cohort (`usq_expr`) to threshold low expression. Genes whose
mean expression falls below the 25th percentile of network genes are flagged
with `is_low_expression = TRUE`.

```{r step4}
annotated_node_metrics <- annotate_low_expressed_genes(
  node_metrics       = annotated_node_metrics,
  expr_data          = usq_expr,
  low_expr_threshold = 0.25
)
```

### Step 5 — ANOVA for subtype-differential expression

Runs a one-way ANOVA for each gene testing whether mean expression differs
across subtypes. Bonferroni correction is applied over all network genes.

```{r step5}
annotated_node_metrics <- run_anova_subtype_expression(
  node_metrics        = annotated_node_metrics,
  expr_data           = sjodahl_2017,
  this_subtype_vector = sjodahl_2017_subtypes,
  sig_threshold       = 0.05
)
```

### Step 6 — Rank subtype expression

For each gene, ranks its mean expression across subtypes (rank 1 = highest
expression), appending `rank_<subtype>` columns.

```{r step6}
annotated_node_metrics <- rank_subtype_expression(
  node_metrics = annotated_node_metrics
)
```

### Step 7 — Add Human Protein Atlas annotations

Joins raw HPA normal-tissue IHC and cancer staining data for the bladder /
urothelial cancer context.

```{r step7}
annotated_node_metrics <- add_hpa_annotations(
  node_metrics       = annotated_node_metrics,
  normal_hpa         = hpa_normal_data,
  cancer_hpa         = hpa_cancer_data,
  this_tissue_normal = "Urinary bladder",
  this_tissue_cancer = "urothelial cancer"
)
```

### Step 8 — Derive HPA expression metrics

Converts the raw HPA fields into interpretable classifications
(`hpa_expression_pattern`, `hpa_passes_filter`, `hpa_therapeutic_target`, etc.).

```{r step8}
annotated_node_metrics <- derive_hpa_metrics(
  node_metrics = annotated_node_metrics
)
```

### Step 9 — Filter network nodes

Retains only genes that are expressed in bladder tissue (normal or cancer),
are not low-expressed, and show statistically significant subtype-differential
expression.

```{r step9}
annotated_node_metrics <- filter_network_nodes(
  node_metrics         = annotated_node_metrics,
  tissue_filter        = "either",
  remove_low_expressed = TRUE,
  filter_anova_sig     = TRUE
)
```

### Step 10 — Recalculate network metrics on the filtered subnetwork

After filtering, centrality values from step 2 reflect the full pre-filter
network. This step rebuilds the graph from the surviving nodes and recomputes
degree, betweenness, closeness, eigenvector centrality, HITS hub score, and
Louvain community membership.

```{r step10}
annotated_node_metrics <- recalculate_network_metrics(
  node_metrics = annotated_node_metrics,
  edge_data    = erbb2_network$edge_data
)
```

### Step 11 — Calculate subtype-specific hub scores

Combines weighted centrality (degree, betweenness, eigenvector) with a
subtype specificity score to produce `<subtype>_hub_score` columns — positive
values indicate high-centrality genes upregulated in that subtype.

```{r step11}
annotated_node_metrics <- calculate_subtype_hub_scores(
  node_metrics = annotated_node_metrics
)
```

---

## Visualisation

### Subtype hub score heatmaps

`plot_subtype_hub_scores()` returns a `ComplexHeatmap` object; use `draw()` to
render it. Inspect upregulated and downregulated hub genes separately.

```{r heatmaps, fig.width = 10, fig.height = 6}
hm_up   <- plot_subtype_hub_scores(annotated_node_metrics, hub_direction = "up")
hm_down <- plot_subtype_hub_scores(annotated_node_metrics, hub_direction = "down")

draw(hm_up)
draw(hm_down)
```

### Network plot

Draw the filtered network as a static `ggraph` plot, colouring nodes by their
origin (seed / pathway / extended set) and labelling all nodes.

```{r network, fig.width = 10, fig.height = 10}
plot_network(
  node_metrics = annotated_node_metrics,
  edge_data    = erbb2_network$edge_data,
  interactive  = FALSE,
  title        = "ERBB2 Network",
  color_by     = "origin",
  label_nodes  = "all"
)
```

### Export to Cytoscape

Write node-attribute and edge TSV files plus a GraphML file for import into
Cytoscape:

```{r cytoscape}
create_cytoscape_objects(
  node_metrics = annotated_node_metrics,
  edge_data    = erbb2_network$edge_data,
  out_dir      = "output/",
  file_prefix  = "erbb2_network"
)
```

---

## Using the `run_netweave()` wrapper

All eleven steps above can be run in a single call:

```{r wrapper}
result <- run_netweave(
  seed_gene          = "ERBB2",
  pathway_name       = "Signaling by ERBB2",
  expr_data          = sjodahl_2017,
  subtype_vector     = sjodahl_2017_subtypes,
  low_expr_data      = usq_expr,
  normal_hpa         = hpa_normal_data,
  cancer_hpa         = hpa_cancer_data,
  this_tissue_normal = "Urinary bladder",
  this_tissue_cancer = "urothelial cancer",
  subtype_order      = c("Uro", "GU", "BaSq", "Mes", "ScNE"),
  string_data_dir    = "data/string_db/"
)

# Access results
final_nodes <- result$node_metrics
edge_data   <- result$edge_data
pathway     <- result$pathway_genes
```

The wrapper accepts the same filter and scoring parameters as the individual
functions; see `?run_netweave` for the full argument list.

---

## Session Info

```{r session}
sessionInfo()
```
